A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm

### 2014-10-30 paint-contour-tree.tex
changed figures
updates to intro

\title{A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm}
\date{}

@@

A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

@@

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
representing the important topological events. An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
Think of the edges as point downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. The correspond to contours (which are themselves manifolds) gaining or losing handles.

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ has $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
Ackermann function.) One cannot overstate the importance of this result, since 
subsequent practical and theoretical work in contour trees uses their ideas. It is natural to think of $N$ as $\Theta(n)$
in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

\begin{theorem} \label{thm:main-alg} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

@@

\begin{theorem} \label{thm:main-lb} For any valid cost $C$, there exists a set $\bF$ of functions over 2-dimensions
with the following properties.
For all $f \in \bF$, $\cost(f) = C$. Any height decision tree that correctly computes
the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
\end{theorem}

@@

Contour trees were (unsurprisingly) used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis gives various
applications of contour trees for data visualization~\cite{c-tmi-04}.

\paragraph{A History of Contour Trees.}
As discussed above, there has been a long series of improved algorithms for computing contour trees.
Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making 
several passes over the input.

Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, 
where $n$ is the number of mesh vertices.  Their algorithm works in 3 simple stages.  
The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
together the join and split trees to get the contour tree.

@@

% \begin{figure}[h]\centering
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
%     \hspace{.5in}
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
%     On the right (in clockwise order from the top) the contour, split and join trees.}
%     \label{fig:sorting}
% \end{figure}
\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.28\linewidth]{figs/balanced2}
    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On bottom (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@

We detail the basic definitions about contour trees. The following is based on Chapter 6 of Carr's thesis \cite{c-tmi-04}.
Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$. 
%All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated \emph{boundary} simplices. 
\Sesh{Some mention of representation and data structures here.}
We use the standard assumption that $f$ is Morse.  In particular, we are assuming no two critical points have the same height, 
and that there are no multi-saddles.

Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$.
We will often omit $f$, and assume that $f$ is encoded in the representation of $\MM$.

\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x \in \MM| f(x) = h\}$.
A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
\end{definition}

Note that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.

@@

\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

This definition allows us to categorize points in $\MM$ depending on the local changes in topology.

\begin{definition} \label{def:points} A point $x \in \MM$ is categorized as follows.
\begin{asparaitem}
	\item Regular: Both up-degree and down-degrees are $1$.
	\item Maximum: Up-degree is $0$.
	\item Minimum: Down-degree is $0$.
	\item Join: Up-degree is strictly greater than $1$.
	\item Split: Down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are called \emph{critical}. Joins and splits are called \emph{saddles}.
Maxima and minima are called \emph{extrema}.
\end{definition}

\begin{remark}
 Note that \Def{deg} can be changed so that the up-degree (resp. down-degree) is the number of $(f(x) + \delta)$-contours (resp. $(f(x) - \delta)$-contours) that intersect $B_\eps(x)$.
 For $2$-manifolds, the two definitions are equivalent, and at each saddle point either two contours join together or split apart, and hence the terminology.  
 In higher dimensions though these definitions differ.  Specifically, this alternative definition only includes points at which the number of contours in a level set change.
 These points are a subset of our defined critical points which also include points at which the topology of the contours change (and hence the term `critical' is still applicable).
 Note that our critical point definition allows one to classify all vertices (according to \Def{points}) simultaneously in linear time, as it is a local neighborhood condition.
\end{remark}

We will enforce a special constraint on boundaries, encapsulated in the following definition.

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
than this value.
\end{definition}

\begin{remark}
Note that by assuming that our manifold is boundary critical, the vertices on a given boundary are either all maxima or all minima.  
Since all such vertices on a boundary component are in the same contour, we will abuse notation and refer to this entire set of vertices as a maxima (resp. minima).
In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
\end{remark}

We use $\cV$ to denote the set of critical points in $\MM$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complexes.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM) = \reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
%
It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$ on $\MM$, and this is known to be a tree for PL functions defined over PL manifolds.

@@

For simplicity we will also make the standard assumption that vertices in the contour 
tree have degree at most $3$, i.e. there are no multi-saddles.
\end{remark}

We use a key theorem from \cite{c-tmi-04} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

@@

@@

>>>>

\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm}
\date{}

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

@@

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical points},
where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
representing the important topological events. An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
Think of the edges pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
Ackermann function.) It is natural to think of $N$ as $\Theta(n)$
in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
pay must $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

\begin{theorem} \label{thm:main-alg} Consider PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and all the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
This is clearly an improvement when the tree is short and fat.
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

@@

\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.
Any height decision tree that correctly computes
the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
\end{theorem}

In other words, it is not possible to improve on \Thm{main-alg} even when restricted to the subset
of inputs of fixed cost.

@@

Contour trees were (unsurprisingly) used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.

There has been a long series of improved algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
One cannot overstate the importance of this result, since it forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.

Contour trees are a special case of Reeb graphs, a general topological representation for real-valued function
on any manifold. We refer the reader to Chapter 6 in~\cite{HaEd10} for more details. Algorithms for computing Reeb graphs
is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.28\linewidth]{figs/balanced2}
    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On bottom (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04},
which provides one of the first rigorous, Morse-theoretic defintions of contour trees. All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear (PL) function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other valued are obtained by interpolation.

We assume that $f$ has distinct values on all vertices, except for boundaries,
as expressed in the following constraint.

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
than this value.
\end{definition}

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in data structures that allow constant-time traversals
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$. 
%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.

\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x| f(x) = h\}$.
A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
\end{definition}

Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.

@@

\begin{definition} \label{def:deg} The Morse up-degree of $x$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The Morse down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

We categorize points depending on the local changes in topology.

\begin{definition} \label{def:points} A point $x$ is categorized as follows.
\begin{asparaitem}
	\item Regular: Both Morse up-degree and down-degrees are $1$.
	\item Maximum: Morse up-degree is $0$.
	\item Minimum: Morse down-degree is $0$.
	\item Morse Join: Morse up-degree is strictly greater than $1$.
	\item Morse Split: Morse down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
Maxima and minima are \emph{extrema}.
\end{definition}

We use $\cV(f)$ to denote the set of critical points.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.
%In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.

\ignore{
The Morse assumption
implies that all critical points have distinct values. A contour that involves a critical
point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
Critical values include these boundary values. By boundary criticality, all contours are themselves
represented as $d-1$-dimensional simplicial complexes.
}

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$, 
known to be a tree for PL functions defined over PL manifolds (Chapter 6 of~\cite{HaEd10}). Alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. We can talk about up-degrees and down-degrees in $\reeb(\MM)$
We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

\subsection{Some technical remarks}

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degrees will have some $>1$ degree in $\reeb(\MM)$.

As Carr points out in his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ can be asociated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explictly placed in $\reeb(\MM)$ (subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
Alternately, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. Our algorithm works for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\ignore{
\begin{remark}
 Note that \Def{deg} can be changed so that the up-degree (resp. down-degree) is the number of $(f(x) + \delta)$-contours (resp. $(f(x) - \delta)$-contours) that intersect $B_\eps(x)$.
 For $2$-manifolds, the two definitions are equivalent, and at each saddle point either two contours join together or split apart, and hence the terminology.  
 In higher dimensions though these definitions differ.  Specifically, this alternative definition only includes points at which the number of contours in a level set change.
 These points are a subset of our defined critical points which also include points at which the topology of the contours change (and hence the term `critical' is still applicable).
 Note that our critical point definition allows one to classify all vertices (according to \Def{points}) simultaneously in linear time, as it is a local neighborhood condition.
\end{remark}

@@

For simplicity we will also make the standard assumption that vertices in the contour 
tree have degree at most $3$, i.e. there are no multi-saddles.
\end{remark}
}

@@

We require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.

@@

Their algorithm works in 3 simple stages.  
The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
together the join and split trees to get the contour tree.

\paragraph{A History of Contour Trees.}

% \begin{figure}[h]\centering
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
%     \hspace{.5in}
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
%     On the right (in clockwise order from the top) the contour, split and join trees.}
%     \label{fig:sorting}
% \end{figure}
